Was Devon Allen Unjustly Disqualified at the 2022 World Track and Field Championships?

### 2023-03-28 Manuscript/manuscript.tex
Major changes to methods and results sections

\jy{Describe you data types; give details on sources; introduce notations for
  both types of data
  present exploratory plots.}
Something that is extremely important to note in the Data is the inconsistency of World Athletics 
in terms of how they penalize false starts.  From 2007-2009, the IAAF (the former name for World 
Athletics, and the governing body for the World Track and Field Champions) instituted a rule change
that allowed one warning false start before a sprinter was disqualified \citep{False-Start}  This 
essentially gave sprinters a second chance and reduced the penalty for a false start.  There were 
18 male and 7 female false starts at the 2007 World Championships, 18 male and 7 female starts at 
the 2009 World Championships. Starting in 2011, the rule was scrapped, and the old policy which 
automatically disqualifies runners who false start was reinstituted.  The effect? At the 2011 World
Championship there were 6 male disqualifications and 4 female disqualifications \citep{False-Start}. 
By returning to the harsher policy and cracking down on false starts, World Athletics had reduced the
number of false starts by two thirds. This was desirable for World Athletics as false starts can 
make already long track meets more tedious, both for athletes and viewers watching the 
television broadcast.  Since 2011, IAAF has instituted the one false start rule
\citep{Pilianidis}.

@@

\subsection{GLMM}

\jy{explain the gamma model in details}

\subsection{Rank-based comparison for Clustered}

\section{Results}

\jy{Presents and discuss the results here}

\section{Concluding Remarks}

\jy{summarize your findings; discuss limitations and possible future works.}

\section{GLMM Model}
\label{sec:GLMM}
Data was copied from the World Athletics website and pasted into an Excel
spreadsheet. The data covers the men's 110 meter hurdles and the women's 100
meter hurdles from 1999 to 2022.  The variable of prediction interest is reaction
time, and the predicting variables are year, stage of competition (heats, 
semifinal, final), total time, and gender.  The data can be found in the appendix:.  
Factors such as name, country, and identification number were
not considered important for the purposes of this research besides the times for
Devon Allen and the other United States athletes mentioned above.  The variable "Stage"
refers to whether the observation occurred during the "heats" (Preliminary round) which
is denoted by "H" in the data, "S" for "semifinals" (Of which there are either 2 or 3 
heats each year), and "F" for finals. The variable "Gender" has levels "M" and "F" 
which stand for "male" and "female" respectively.  The variable "Batch" takes a 
numerical value and every heat of every round from 1999 to 2022 was assigned it's
own number.  This was added because of suspected issues that reaction time may be
correlated with other runners reacting quickly (one runner reacting quickly may
result in other runners reacting quickly).  There is no statistical signficance to
how the batches were labeled, they were ordered chronologically by meet but
reverse chronologically by year (Batch 1 is a preliminary Heat of 2022 and Batch
111 corresponds to the 1999 finals). The data that looked at in the paper
is best visualized using a scatterplot and box plot. Figure~\ref{fig:DataBoxPlot}
below shows the box plot data.

One of the most interesting things found was that the reaction times from the USA
athletes in the World Championship 110 meter hurdles, was that they were significantly
faster than at the USA Track Championships.  From June 23-26 2022, USA held its Track
and Field Championship to decide who to send to the World Championships at Hayward 
Field, the exact same venue as used for the World Track and Field Championship a month 
later.  Thus, a baseline is able to be established for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all of the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field  Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teamates.  This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  This table below highlights the difference in reaction times for
Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. "USA" denotes
the USA Track and Field Championships, "World" denotes the World Track and Field
Championships, "N/A" was used for any athlete that did not compete and thus
did not register a reaction time.  All numbers listed in the table are the reaction
time for the respective athletes in seconds.

@@

All Men final's Data
\begin{center}
  \begin{tabular}{|c | c | c | c | c | c | c |} 
   \hline\hline
   Model & AIC & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline
   linear one & -472.7 & 238.3 & 0.1484 & 0.0019 & N/A & N/A \\ 
   \hline
   linear year & -468.7 & 237.4 & 0.1481 & 0.0030 & 7.762e-05 & 2.457e-04 \\
   \hline
   men finals one gamma & -477.3 & 240.7 & 6.738 & 0.0844 & N/A & N/A \\
   \hline
   men finals year gamma & -491.1 & 248.5 & 6.8121 & 0.1895 & 0.1148 & 0.0112 \\
   \hline
   men finals one gamma log & -477.3 & 240.7 & -1.908 & 0.0125 & N/A & N/A \\
   \hline
   men finals year gamma log & -491.0 & 248.5 & -1.914 & 0.0280 & 0.0025 & 0.0112 \\ [0.5ex]
   \hline
  \end{tabular}
\end{center}

@@

All Men's semi finals and finals data
\begin{center}
  \begin{tabular}{|c | c | c | c | c | c | c |} 
   \hline
   Model & AIC & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline\hline
   linear one & -1667 & 835.5 & 0.1529 & 0.0014 & N/A & N/A \\
   \hline
   linear year & -1713 & 859.3 & 0.1541 & 0.0040 & 1.719e-04 & 5.926e-04 \\ 
   \hline
   gamma one & -1756 & 880.2 & 6.554 & 0.0598 & N/A & N/A \\
   \hline
   gamma year & -1848 & 926.9 & 6.553 & 0.2269 & 0.1305 & 0.0234 \\
   \hline
   gamma one log & -1756 & 880.2 & -1.880 & 0.0091 & N/A & N/A \\
   \hline
   gamma year log & -1848 & 927.0 & -1.869 & 0.0348 & 0.0031 & 0.0235 \\ [0.5ex]
   \hline
  \end{tabular}
  \end{center}

@@

AIC with and without random effect, Gamma and log model on all semifinal and finals data
\begin{center}
  \begin{tabular}{|c | c|}
    \hline
    Random Effect? & AIC \\
    No & -1756 \\
    Yes & -1848 \\
    \hline
  \end{tabular}
\end{center}

Addition of 2022 on Gamma and log model on semifinal and finals data
\begin{center}
  \begin{tabular}{|c | c|}
    \hline
    Dataset & venue effect Standard Error \\
    1999-2019 & 0.0495 \\
    1999-2022 & 0.0554 \\
    \hline
  \end{tabular}
\end{center}

All summary statistics for Gamma and log model on all semifinal and finals data
\begin{center}
  \begin{tabular}{| c | c | c | c | c |}
    \hline
    AIC & BIC & logLik & Random Effect standard dev & Residual standard dev \\
    -1847 & -1836 & 927.0 & 0.0031 & 0.0235 \\
    \hline
  \end{tabular}
\end{center}

\section{GLMM Gamma Model}

In terms of the data that was used, it is worth noting that the standard error
of the intercept was higher for all 6 comparisons of the data: finals data and 
pooled data for the linear model, gamma model, gamma and log model for when
2022 was included versus when it was not.  This shows that the addition
of 2022 to the linear model increased the model's variation and that the addition
of 2022 resulted in a worse model.  Although this is more of a qualitative analysis,
this is further evidence that 2022 was a significant year in the role it played
in altering the gamma model.

\section{Further Analysis}
, but that is not necessarily the results that were found.  The R package
"clusrank" was used to perform a clustered Wilcoxon rank sum test.  This was
chosen over a typical t test so as to avoid any assumptions about the distribution
of population means.  The clusrank package allows clusters to be formed, which
in this case are for each athlete.  The data containing each athlete's name,
nationality, and reaction times for the 2019 and 2022 World Championship can be
found in the csv file labeled "2019vs2022".  This data was transformed to "Clusrank2019vs2022"
so that it better fits the clusrank package.  Each athlete was given a non-meaningful
ID (An ID of 1 does not indicate that this athlete is better or worse than any
other athlete).  Additionally, the "Year" column is used to determine what year
the time was registered with "1" indicating 2019 and "2" indicating 2022.

%Citations needed
Using the "clusWilcox.test" function, a grouped and clustered Wilcoxon rank sum
test was performed wtih the cluster set to be the Athlete and group set to Year.
The method was set to be "ds" which is used for the Data and Satten method, which
is preferable when the cluster sizes are informative.  We do not care that some
athletes made it further in either competition than another or that it should
not matter how elite an athlete is at the 110 meter hurdles once we get to the
national international competition level.  Cluster sizes vary from 2 observations
to 6.

There is not a reasonable explanation for this.  It is inconceivable that so many
athletes would improve such a significant amount in such a small amount of time.
6 countries are represented in this comparison, which drastically decreases the
chances of this issue being caused by one specific country.  The United States
athlete's improvement in times was discussed earlier in the paper, but this data
%Use a reference in this part
shows that on average there was improvement for British, Polish, French, Brazilian,
and Spanish athletes.

\section{Conclusion}
\label{sec:conclusion}
It is worth noting that in the semi-finals of the World Track and
Field Championships; Allen's reaction time was 0.101 which is only 0.001 above
the legal limit.  What this suggests is that Allen may have strong reflexes
and be able to react extremely well to the sound of the gun.  It seems unlikely that
Allen would be able to correctly predict the gun with such precision two times in a row,
which is the exact reason that the 0.10 second reaction time disqualification barrier was
imposed in the first place.  But Allen showed that at the Championship he was able to
react extremely fast two times, which may suggest skill rather than luck. Devon Allen is a very good athlete, as evidenced by his football career at the University of Oregon
and making it onto the Philadelphia Eagles practice squad this past year \citep{Hurley}.
Thus the 0.099 reaction time may not have been a product of Devon Allen
predicting the start of the race but rather a combination of a quick reaction 
and a possibly faulty sensor.

The figure below plots the average reaction time by year for the pooled men's data since
1999.  The circle in the bottom right is for 2022, which is significantly lower than the
data points for the previous three years, and is substially lower than the data points
for 2011 and 2013.  This graph supports the conclusion that the timing system from 2022
had some sort of error that caused the reaction times to be so much lower.  Assuming there
was not an issue with the system, the large variation is confusing because there is little explanation
as to why there is so much of a dropoff in reaction time from 2013 to 2015 or the spike
from 2001 to 2003.  The variation from year to year is slightly concerning in the
sense that if there is a strong venue effect then there is strong variation between
years.  This is not desirable and what this could possibly suggest is that
there is a lot of randomness in how athletes react or that there is at least one
other factor that affects reaction time. If the timing systems are succeptible to
external factors that affect how athletes react, then this is something that 
IAAF may want to consider when they select cities to host an important event 
such as the Track and Field World Championship. Additionally, Seiko may want to 
do more research on their timing systems to see if climate factors can impact
the precision of the accuracy of the equipment.

To try and approximate the distribution of the reaction times, R code was written
to try and estimate the probabilites of observing a reaction time as extreme as
0.1 seconds.At a size of n=12,000,000, the calculated probability of a reaction
time being below 0.1 seconds was as follows for the various different data sets:

\begin{center}
  \begin{tabular}{||c|c||} 
   \hline
   Data Set & P RxnTime 0.1 \\ 
   \hline\hline
   all finals gamma log & 0.000864 \\ 
   \hline
   old finals gamma log & 0.000429 \\
   \hline
   all pooled gamma log & 0.006847 \\
   \hline
   old pooled gamma log & 0.005749 \\
  \end{tabular}
  \end{center}

>>>>

\jy{Describe you data types; give details on sources; introduce notations for
  both types of data
  present exploratory plots.}

Something that is extremely important to note in the Data is the inconsistency of World Athletics 
in terms of how they penalize false starts.  From 2007-2009, the IAAF (the former name for World 
Athletics, and the governing body for the World Track and Field Champions) instituted a rule change
that allowed one warning false start before a sprinter was disqualified \citep{False-Start}  This 
essentially gave sprinters a second chance and reduced the penalty for a false start.  There were 
18 male and 7 female false starts at the 2007 World Championships, 18 male and 7 female starts at 
the 2009 World Championships. Starting in 2011, the rule was scrapped, and the old policy which 
automatically disqualifies runners who false start was reinstituted.  The effect? At the 2011 World
Championship there were 6 male disqualifications and 4 female disqualifications \citep{False-Start}. 
By returning to the harsher policy and cracking down on false starts, World Athletics had reduced the
number of false starts by two thirds. This was desirable for World Athletics as false starts can 
make already long track meets more tedious, both for athletes and viewers watching the 
television broadcast.  Since 2011, IAAF has instituted the one false start rule
\citep{Pilianidis}.

Data was copied from the World Athletics website and pasted into an Excel
spreadsheet. The data covers the men's 110 meter hurdles and the women's 100
meter hurdles from 1999 to 2022.  The variable of prediction interest is reaction
time, and the predicting variables are year, stage of competition (heats, 
semifinal, final), total time, and gender.  The data can be found in the appendix:.  
Factors such as name, country, and identification number were
not considered important for the purposes of this research besides the times for
Devon Allen and the other United States athletes mentioned above.  The variable "Stage"
refers to whether the observation occurred during the "heats" (Preliminary round) which
is denoted by "H" in the data, "S" for "semifinals" (Of which there are either 2 or 3 
heats each year), and "F" for finals. The variable "Gender" has levels "M" and "F" 
which stand for "male" and "female" respectively.  The variable "Batch" takes a 
numerical value and every heat of every round from 1999 to 2022 was assigned it's
own number.  This was added because of suspected issues that reaction time may be
correlated with other runners reacting quickly (one runner reacting quickly may
result in other runners reacting quickly).  There is no statistical signficance to
how the batches were labeled, they were ordered chronologically by meet but
reverse chronologically by year (Batch 1 is a preliminary Heat of 2022 and Batch
111 corresponds to the 1999 finals). The data that looked at in the paper
is best visualized using a box plot. Figure~\ref{fig:DataBoxPlot}.  Included are
box plots of the reaction times of the men's semi finals and finals pooled
together and just the finals.

\begin{figure}[h]
  \centering 
  \includegraphics{DataBoxPlot}
  \label{fig:DataBoxPlot}
\end{figure}

\begin{figure}[h]
  \centering 
  \includegraphics{BoxplotFinals}
  \label{fig:BoxplotFinals}
\end{figure}

@@

\subsection{GLMM}
A gamma mixed linear effects model was chosen as the best model to model the data.
Many models were tried and compared to determine what the strongest model.  For
a fixed data set such as the men's pooled data including 2022, the models with
the venue effect included performed better than the models without the venue
effect for each set of comparable models; as indicated by the lower AIC and
higher log likelihood.  Thus, the venue effect is significant and should be
included in the model.  It can also be established that a gamma distribution
better models the data than linear data from comparing the linear model with
venue effect and them gamma model with a venue effect.  The final decision that
was made in model selection was to include link = log in the model.  Despite
%Does this part need to be changed?
doing very little to alter the AIC or log likelihood, the log scale was included
in order to try and better deal with high outliers. %Include Residuals plot?

All Men's semi finals and finals data
\begin{center}
  \begin{tabular}{|c  c  c  c  c  c  c |}
   \hline 
   Model & AIC & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline\hline
   linear one & -1667 & 835.5 & 0.1529 & 0.0014 & N/A & N/A \\
   \hline
   linear year & -1713 & 859.3 & 0.1541 & 0.0040 & 1.719e-04 & 5.926e-04 \\ 
   \hline
   gamma one & -1756 & 880.2 & 6.554 & 0.0598 & N/A & N/A \\
   \hline
   gamma year & -1848 & 926.9 & 6.553 & 0.2269 & 0.1305 & 0.0234 \\
   \hline
   gamma one log & -1756 & 880.2 & -1.880 & 0.0091 & N/A & N/A \\
   \hline
   gamma year log & -1848 & 927.0 & -1.869 & 0.0348 & 0.0031 & 0.0235 \\ [0.5ex]
   \hline
  \end{tabular}
  \end{center}

\subsection{Rank-based comparison for Clustered}
This section was motivated from a qualitative analysis that started by looking
at how United States athletes reacted at the 2022 United States Track and Field
Championships. From June 23-26 2022, USA held its Track
and Field Championship to decide who to send to the World Championships at Hayward 
Field, the exact same venue as used for the World Track and Field Championship a month 
later.  Thus, a baseline is able to be established for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all of the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field  Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teamates.  This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  This table below highlights the difference in reaction times for
Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. "USA" denotes
the USA Track and Field Championships, "World" denotes the World Track and Field
Championships, "N/A" was used for any athlete that did not compete and thus
did not register a reaction time.  All numbers listed in the table are the reaction
time for the respective athletes in seconds.

@@

This dataset is quite small to perform any formal statistical analysis on, but
the data was expanded to include data for athletes who competed at national
competitions (many of which were held from May-July of 2022).  Once the data
was collected, each athlete was designated to be a cluster and the stage of
competition (national or international) was specified to be the group.  Next,
a Clustered Wilcoxon rank sum test was performed using the clusrank R package
\citep{Clusrank}.  Using a clustered rank sum test was chosen over a typical t 
test so as to avoid any assumptions about the distribution of population means.

%Citations needed
Using the "clusWilcox.test" function, a grouped and clustered Wilcoxon rank sum
test was performed wtih the cluster set to be the Athlete and group set to Year.
The method was set to be "ds" which is used for the Data and Satten method, which
is preferable when the cluster sizes are informative \citep{Clusrank}.  We do not care that some
athletes made it further in either competition than another or that it should
not matter how elite an athlete is at the 110 meter hurdles once we get to the
national international competition level.  Cluster sizes vary from 2 observations
to 6.

\section{Results}
\jy{summarize your findings; discuss limitations and possible future works.}

To try and approximate the distribution of the reaction times, R code was written
to try and estimate the probabilities of observing a reaction time as extreme as
0.1 seconds. At a size of n=12,000,000, the calculated probability of a reaction
time being below 0.1 seconds was as follows for the various different data sets:

\begin{center}
  \begin{tabular}{|c c|} 
   \hline
   Data Set & Probability RxnTime < 0.1 \\ 
   \hline\hline
   all finals gamma log & 0.000864 \\ 
   \hline
   old finals gamma log & 0.000429 \\
   \hline
   all pooled gamma log & 0.006847 \\
   \hline
   old pooled gamma log & 0.005749 \\
  \end{tabular}
  \end{center}

It is worth noting that changing the reaction time barrier from 0.1 seconds to
0.08 seconds drastically reduces the chances of observing a reaction time that
would break the barrier.  For the dataset of all finals data, the probability
associated with observing a reaction time below 0.08 seconds was $6.67\cdot10{-7}$.
If IAAF were to change the reaction time the way that \citet{Komi}'s paper
suggests, the probability would be so much lower that athletes would likely not
have a case to make when arguing against a disqualification.

The p values produced by the clustered rank sum tests were 0.000155 for the
same athletes competing in the same year, and 0.001464 for the same athletes in
different years.  To put it simply, there is not a reasonable explanation for 
this.  
It is inconceivable that so many athletes would improve such a significant
amount in such a small amount of time. 6 countries are represented in this 
comparison, which drastically decreases the chances of this issue being caused by
one specific country.  The United States athlete's improvement in times was
discussed earlier in the paper, but this data shows that on average there was
%Use a reference in this part
improvement for British, Polish, French, Brazilian, and Spanish athletes.

It is worth noting that in the semi-finals of the World Track and
Field Championships; Allen's reaction time was 0.101 which is only 0.001 above
the legal limit.  What this suggests is that Allen may have strong reflexes
and be able to react extremely well to the sound of the gun.  It seems unlikely that
Allen would be able to correctly predict the gun with such precision two times in a row,
which is the exact reason that the 0.10 second reaction time disqualification barrier was
imposed in the first place.  But Allen showed that at the Championship he was able to
react extremely fast two times, which may suggest skill rather than luck. Devon Allen is a very good athlete, as evidenced by his football career at the University of Oregon
and making it onto the Philadelphia Eagles practice squad this past year \citep{Hurley}.
Thus the 0.099 reaction time may not have been a product of Devon Allen
predicting the start of the race but rather a combination of a quick reaction 
and a possibly faulty sensor.

Here are important statistics grouped by both model and data set:
All Men final's Data
\begin{center}
  \begin{tabular}{|c | c | c | c | c | c | c |} 
   \hline\hline
   Model & AIC & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline
   linear one & -472.7 & 238.3 & 0.1484 & 0.0019 & N/A & N/A \\ 
   \hline
   linear year & -468.7 & 237.4 & 0.1481 & 0.0030 & 7.762e-05 & 2.457e-04 \\
   \hline
   men finals one gamma & -477.3 & 240.7 & 6.738 & 0.0844 & N/A & N/A \\
   \hline
   men finals year gamma & -491.1 & 248.5 & 6.8121 & 0.1895 & 0.1148 & 0.0112 \\
   \hline
   men finals one gamma log & -477.3 & 240.7 & -1.908 & 0.0125 & N/A & N/A \\
   \hline
   men finals year gamma log & -491.0 & 248.5 & -1.914 & 0.0280 & 0.0025 & 0.0112 \\ [0.5ex]
   \hline
  \end{tabular}
\end{center}

@@

All Men's semifinal and finals data
\begin{center}
  \begin{tabular}{||c | c c c | c c c||} 
   \hline
   Athlete & USA H & USA S & USA F & World H & World S & World F \\ [0.5ex] 
   \hline\hline
   Devon Allen & 0.201 & 0.153 & 0.160 & 0.123 & 0.101 & 0.099 \\ 
   \hline
   Trey Cunningham & 0.186 & 0.185 & 0.182 & 0.115 & 0.120 & 0.109 \\
   \hline
   Grant Holloway & 0.192 & 0.190 & N/A & 0.147 & 0.128 & 0.124 \\
   \hline
   Daniel Roberts & 0.181 & 0.200 & 0.183 & 0.179 & N/A & N/A \\ [0.5ex]
   \hline
  \end{tabular}
  \end{center}

@@
